---
title: "Sentinel Dashboard Workflow -> Package-Ready Functions"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Purpose

This document translates the Shiny workflow into a function-first API so an enthusiastic coder can run the same logic from scripts or a future R package.

Design goals:

- one function per workflow action
- explicit arguments and return values
- simpler internals than app reactivity
- clear split between **Upload** and **Retrieve/Create** paths

You can later move each function into `R/*.R` and keep the same signatures.

```{r}
#| label: setup-libraries
#| eval: false
library(sf)
library(terra)
library(raster)
library(rstac)
library(dplyr)
library(mapview)
library(mapedit)
library(leaflet)
library(jsonlite)
```

# 1) Boundary Tab

## 1.1 Upload/Read Boundary

```{r}
#| label: fn-read-boundary
#| eval: false
#' Read a boundary file and enforce a single polygon AOI.
#'
#' @param path Path to .geojson/.kml/.kmz/.zip boundary file.
#' @param target_crs Output CRS for AOI (default EPSG:4326).
#' @return sf polygon object with one feature.
read_boundary <- function(path, target_crs = 4326) {
  ext <- tolower(tools::file_ext(path))
  tmp_dir <- tempfile("boundary_")
  dir.create(tmp_dir, showWarnings = FALSE, recursive = TRUE)
  on.exit(unlink(tmp_dir, recursive = TRUE), add = TRUE)

  if (ext == "zip" || ext == "kmz") {
    unzip(path, exdir = tmp_dir)
    candidates <- list.files(
      tmp_dir,
      pattern = "\\.(geojson|kml|shp)$",
      full.names = TRUE,
      recursive = TRUE
    )
    if (!length(candidates)) stop("No supported boundary file found inside archive.")
    path <- candidates[[1]]
  }

  aoi <- sf::st_read(path, quiet = TRUE)
  if (nrow(aoi) < 1) stop("Boundary file is empty.")
  aoi <- aoi[1, , drop = FALSE]
  aoi <- sf::st_make_valid(aoi)
  sf::st_transform(aoi, target_crs)
}

```

## 1.2 Create Boundary (No Upload Needed)

```{r}
#| label: fn-create-boundary
#| eval: false
#' Create an AOI polygon directly from coordinates.
#'
#' @param xmin,ymin,xmax,ymax Bounding box coordinates in longitude/latitude.
#' @param target_crs Output CRS.
#' @return sf polygon AOI.
create_boundary_bbox <- function(xmin, ymin, xmax, ymax, target_crs = 4326) {
  bb <- sf::st_bbox(c(xmin = xmin, ymin = ymin, xmax = xmax, ymax = ymax), crs = sf::st_crs(4326))
  aoi <- sf::st_as_sfc(bb)
  aoi <- sf::st_sf(id = 1, geometry = aoi)
  sf::st_transform(aoi, target_crs)
}
```

## 1.3 Create Boundary Visually (Map + Current Location)

```{r}
#| label: fn-create-boundary-visual
#| eval: false
#' Guess current location from IP (optional convenience helper).
#'
#' @return named numeric vector c(lon, lat).
guess_current_location <- function() {
  x <- jsonlite::fromJSON("https://ipapi.co/json/")
  c(lon = as.numeric(x$longitude), lat = as.numeric(x$latitude))
}

#' Draw AOI polygon interactively on a map.
#'
#' @param start_lon,start_lat Initial map center.
#' @param zoom Initial map zoom level.
#' @param target_crs Output CRS for AOI.
#' @return sf polygon AOI drawn by user.
create_boundary_interactive <- function(start_lon = NULL, start_lat = NULL, zoom = 12, target_crs = 4326) {
  if (is.null(start_lon) || is.null(start_lat)) {
    loc <- tryCatch(guess_current_location(), error = function(e) c(lon = -83.3576, lat = 33.9519))
    start_lon <- loc[["lon"]]
    start_lat <- loc[["lat"]]
  }

  m <- leaflet::leaflet() %>%
    leaflet::addProviderTiles("Esri.WorldImagery") %>%
    leaflet::setView(lng = start_lon, lat = start_lat, zoom = zoom)

  # User draws a polygon then saves in Viewer pane.
  drawn <- mapedit::editMap(m)
  aoi <- drawn$finished
  if (is.null(aoi) || nrow(aoi) < 1) stop("No polygon drawn. Draw AOI and save from map editor.")
  aoi <- aoi[1, , drop = FALSE]
  aoi <- sf::st_make_valid(aoi)
  sf::st_transform(aoi, target_crs)
}


bbox <- create_boundary_interactive()
```

## 1.4 Validate Boundary Size

```{r}
#| label: fn-boundary-area-check
#| eval: false
#' Compute AOI area summary for warnings.
#'
#' @param aoi sf polygon.
#' @return list(area_km2, level) where level in c("ok","caution","warning").
boundary_area_check <- function(aoi) {
  aoi_m <- sf::st_transform(aoi, 3857)
  km2 <- as.numeric(sf::st_area(aoi_m)) / 1e6
  level <- if (km2 > 100) "warning" else if (km2 > 50) "caution" else "ok"
  list(area_km2 = km2, level = level)
}

boundary_area_check(bbox)
```

# 2) Imagery Tab

## 2.1 Retrieve Path: Sentinel (Search)

```{r}
#| label: fn-search-sentinel
#| eval: false
#' Search Sentinel-2 scenes from Planetary Computer STAC.
#'
#' @param aoi sf polygon in any CRS.
#' @param date_start Start date "YYYY-MM-DD".
#' @param date_end End date "YYYY-MM-DD".
#' @param cloud_max Max cloud cover percent.
#' @param limit Max returned items before local filtering.
#' @return list of STAC features.
search_sentinel <- function(aoi, date_start, date_end, cloud_max = 10, limit = 20) {
  aoi_ll <- sf::st_transform(aoi, 4326)
  bb <- sf::st_bbox(aoi_ll)
  bbox <- c(bb["xmin"], bb["ymin"], bb["xmax"], bb["ymax"])
  datetime <- paste0(date_start, "/", date_end)

  s <- rstac::stac("https://planetarycomputer.microsoft.com/api/stac/v1/")
  items <- s %>%
    rstac::stac_search(
      collections = "sentinel-2-l2a",
      bbox = bbox,
      datetime = datetime,
      limit = limit
    ) %>%
    rstac::get_request()

  feats <- items$features %||% list()
  cc <- vapply(feats, function(f) suppressWarnings(as.numeric(f$properties$`eo:cloud_cover`)), numeric(1))
  keep <- is.na(cc) | cc <= cloud_max
  feats[keep]
}

senti <- search_sentinel(bbox, "2025-06-01", "2026-03-15")

senti[[1]]$links
```

## 2.2 Retrieve Path: Sentinel (Single Scene)

```{r}
#| label: fn-retrieve-sentinel-single
#| eval: false
#' Build a Sentinel stack for one selected scene.
#'
#' @param feature One STAC feature from search_sentinel().
#' @param aoi sf polygon AOI.
#' @param out_tif Optional output path. If NULL, temp file is used.
#' @param target_crs Output CRS string (e.g., "EPSG:4326").
#' @return list(path, rast, names).
retrieve_sentinel_single <- function(feature, aoi, out_tif = NULL, target_crs = "EPSG:4326") {
  if (is.null(out_tif)) out_tif <- tempfile(fileext = ".tif")

  signed <- rstac::items_sign(
    list(type = "FeatureCollection", features = list(feature)),
    sign_fn = rstac::sign_planetary_computer()
  )
  f <- signed$features[[1]]

  get_url <- function(asset_name) paste0("/vsicurl/", f$assets[[asset_name]]$href)
  b02 <- terra::rast(get_url("B02"))
  b03 <- terra::rast(get_url("B03"))
  b04 <- terra::rast(get_url("B04"))
  b05 <- terra::rast(get_url("B05"))
  b08 <- terra::rast(get_url("B08"))

  aoi_v <- terra::vect(sf::st_transform(aoi, terra::crs(b02)))
  clip <- function(r) terra::mask(terra::crop(r, aoi_v), aoi_v)

  blue <- clip(b02); green <- clip(b03); red <- clip(b04); re <- clip(b05); nir <- clip(b08)
  re <- terra::resample(re, nir, method = "bilinear")

  stack <- terra::c(blue, green, red, re, nir)
  names(stack) <- c("Blue", "Green", "Red", "RedEdge", "NIR")
  stack <- terra::project(stack, target_crs)

  terra::writeRaster(stack, out_tif, overwrite = TRUE, datatype = "FLT4S")
  list(path = out_tif, rast = stack, names = names(stack))
}

```

## 2.3 Retrieve Path: Sentinel (Median Composite)

```{r}
#| label: fn-retrieve-sentinel-median
#| eval: false
#' Build a median Sentinel stack from multiple scenes.
#'
#' @param features STAC feature list from search_sentinel().
#' @param aoi sf polygon AOI.
#' @param n_use Number of scenes to include.
#' @param out_tif Optional output path.
#' @param target_crs Output CRS.
#' @return list(path, rast, names).
retrieve_sentinel_median <- function(features, aoi, n_use = 8, out_tif = NULL, target_crs = "EPSG:4326") {
  if (length(features) < 2) stop("Need at least 2 scenes for median.")
  n_use <- max(2L, min(as.integer(n_use), length(features)))
  sub_feats <- features[seq_len(n_use)]

  stacks <- lapply(sub_feats, function(f) retrieve_sentinel_single(f, aoi, target_crs = target_crs)$rast)
  ref <- stacks[[1]]
  if (length(stacks) > 1L) {
    for (i in 2:length(stacks)) stacks[[i]] <- terra::resample(stacks[[i]], ref, method = "bilinear")
  }

  med <- vector("list", 5L)
  for (i in seq_len(5L)) {
    rs <- terra::rast(lapply(stacks, function(s) s[[i]]))
    med[[i]] <- terra::app(rs, fun = function(z) stats::median(z, na.rm = TRUE))
  }

  out <- terra::rast(med)
  names(out) <- paste0(c("Blue", "Green", "Red", "RedEdge", "NIR"), "_median")
  if (is.null(out_tif)) out_tif <- tempfile(fileext = ".tif")
  terra::writeRaster(out, out_tif, overwrite = TRUE, datatype = "FLT4S")
  list(path = out_tif, rast = out, names = names(out))
}

retrieve_sentinel_median(senti, bbox)
```

## 2.4 Retrieve Path: Planet

```{r}
#| label: fn-retrieve-planet-skeleton
#| eval: false
#' Retrieve Planet imagery (skeleton wrapper).
#'
#' @param aoi sf AOI polygon.
#' @param api_key Planet API key.
#' @param date_start,date_end Date range.
#' @param cloud_max Max cloud cover.
#' @return path to processed raster/zip outputs.
retrieve_planet <- function(aoi, api_key, date_start, date_end, cloud_max = 10) {
  stop("Implement by reusing your app's Planet query/order/download steps in one pure function.")
}
```

## 2.5 Upload Path: Imagery Stack / Individual Bands / Precomputed VI

```{r}
#| label: fn-upload-ms-stack
#| eval: false
#' Read user-uploaded multispectral stack.
#'
#' @param tif_path Path to stack TIFF.
#' @param band_names Optional names vector.
#' @return raster::brick
upload_ms_stack <- function(tif_path, band_names = NULL) {
  r <- raster::brick(tif_path)
  if (!is.null(band_names)) names(r) <- band_names
  r
}
```

```{r}
#| label: fn-stack-individual-bands
#| eval: false
#' Stack individual band TIFFs into a multispectral brick.
#'
#' @param blue,green,red,rededge,nir Paths to TIFF files.
#' @return raster::brick with standard names.
stack_individual_bands <- function(blue, green, red, rededge, nir) {
  b <- raster::raster(blue)
  g <- raster::raster(green)
  r <- raster::raster(red)
  re <- raster::raster(rededge)
  n <- raster::raster(nir)
  out <- raster::brick(b, g, r, re, n)
  names(out) <- c("Blue", "Green", "Red", "RedEdge", "NIR")
  out
}
```

```{r}
#| label: fn-add-uploaded-vi
#| eval: false
#' Add uploaded vegetation index layer to a named list.
#'
#' @param vi_list Named list of RasterLayer/SpatRaster objects.
#' @param vi_name New layer name (e.g., "NDVI").
#' @param vi_tif Path to TIFF.
#' @return Updated named list.
add_uploaded_vi <- function(vi_list, vi_name, vi_tif) {
  vi_list[[vi_name]] <- raster::raster(vi_tif)
  vi_list
}
```

# 3) Elevation Tab

## 3.1 Retrieve Path: Elevation Download

```{r}
#| label: fn-download-elevation
#| eval: false
#' Download DEM for AOI and project to target CRS.
#'
#' @param aoi sf AOI polygon.
#' @param z Elevation zoom level/resolution proxy.
#' @param target_crs Output CRS.
#' @return terra SpatRaster DEM.
download_elevation <- function(aoi, z = 12, target_crs = "EPSG:4326") {
  dem <- elevatr::get_elev_raster(sf::st_transform(aoi, 4326), z = z, clip = "locations")
  dem <- terra::rast(dem)
  terra::project(dem, target_crs)
}
```

## 3.2 Upload Path: DEM and Derivatives

```{r}
#| label: fn-upload-elevation-layer
#| eval: false
#' Read an uploaded elevation/terrain TIFF.
#'
#' @param tif_path Path to .tif/.tiff.
#' @return raster::raster object.
upload_elevation_layer <- function(tif_path) {
  raster::raster(tif_path)
}
```

# 4) Other Covariates Tab

## 4.1 Upload Path: Arbitrary Raster Layer

```{r}
#| label: fn-add-other-raster
#| eval: false
#' Add an arbitrary uploaded raster as a named layer.
#'
#' @param layer_list Existing named list of rasters.
#' @param name Layer name.
#' @param tif_path Raster file path.
#' @return Updated named list.
add_other_raster <- function(layer_list, name, tif_path) {
  layer_list[[name]] <- raster::raster(tif_path)
  layer_list
}
```

# 5) Sampling Tab

## 5.1 Manual Sampling

```{r}
#| label: fn-manual-points
#| eval: false
#' Build manual point sf from coordinates.
#'
#' @param df Data frame with lon/lat columns.
#' @param lon_col,lat_col Column names.
#' @param crs CRS for created points (default EPSG:4326).
#' @return sf POINT object.
build_manual_points <- function(df, lon_col = "lon", lat_col = "lat", crs = 4326) {
  sf::st_as_sf(df, coords = c(lon_col, lat_col), crs = crs, remove = FALSE)
}
```

```{r}
#| label: fn-manual-sampling-visual
#| eval: false
#' Draw manual sample points directly on an interactive map.
#'
#' @param aoi sf AOI polygon.
#' @param target_crs CRS for returned points.
#' @return sf POINT object.
manual_sampling_interactive <- function(aoi, target_crs = 4326) {
  aoi_ll <- sf::st_transform(aoi, 4326)
  m <- mapview::mapview(aoi_ll, alpha.regions = 0.15, lwd = 2)
  drawn <- mapedit::editMap(m@map)
  pts <- drawn$finished
  if (is.null(pts) || nrow(pts) < 1) stop("No points drawn. Add points and save.")
  pts <- sf::st_cast(pts, "POINT", warn = FALSE)
  sf::st_transform(pts, target_crs)
}
```

## 5.2 Generate Samples (Random / cLHS / Grid)

```{r}
#| label: fn-generate-samples-skeleton
#| eval: false
#' Generate sample points from covariates and AOI.
#'
#' @param aoi sf AOI polygon.
#' @param covariates RasterStack/SpatRaster.
#' @param method "random", "clhs", or "grid".
#' @param n_points Number of points (for random/clhs).
#' @return sf POINT object.
generate_samples <- function(aoi, covariates, method = c("random", "clhs", "grid"), n_points = 30) {
  method <- match.arg(method)
  aoi <- sf::st_transform(aoi, 4326)
  cov <- terra::rast(covariates)
  aoi_v <- terra::vect(aoi)

  # Align AOI and covariates once.
  if (!identical(terra::crs(cov), terra::crs(aoi_v))) {
    aoi_v <- terra::project(aoi_v, terra::crs(cov))
  }
  cov <- terra::crop(cov, aoi_v, snap = "out")
  cov <- terra::mask(cov, aoi_v)

  if (method == "random") {
    pts <- terra::spatSample(cov[[1]], size = n_points, method = "random", na.rm = TRUE, as.points = TRUE)
    return(sf::st_as_sf(pts))
  }

  if (method == "grid") {
    bb <- sf::st_bbox(sf::st_as_sf(aoi_v))
    side <- ceiling(sqrt(max(1, n_points)))
    xs <- seq(bb["xmin"], bb["xmax"], length.out = side)
    ys <- seq(bb["ymin"], bb["ymax"], length.out = side)
    grd <- expand.grid(x = xs, y = ys)
    pts <- sf::st_as_sf(grd, coords = c("x", "y"), crs = sf::st_crs(sf::st_as_sf(aoi_v)))
    keep <- sf::st_within(pts, sf::st_as_sf(aoi_v), sparse = FALSE)[, 1]
    return(pts[keep, , drop = FALSE])
  }

  # cLHS path
  if (!requireNamespace("clhs", quietly = TRUE)) {
    stop("Package 'clhs' is required for method='clhs'. Install with install.packages('clhs').")
  }
  vals <- terra::as.data.frame(cov, xy = TRUE, na.rm = TRUE)
  if (nrow(vals) < n_points) stop("Not enough non-NA covariate cells for requested cLHS sample size.")
  cl <- clhs::clhs(vals[, setdiff(names(vals), c("x", "y"))], size = n_points, progress = FALSE, simple = FALSE)
  sel <- vals[cl$index_samples, c("x", "y"), drop = FALSE]
  sf::st_as_sf(sel, coords = c("x", "y"), crs = sf::st_crs(sf::st_as_sf(aoi_v)))
}
```

# 6) Shared Utilities

## 6.1 Compute Vegetation Indices

```{r}
#| label: fn-calc-vis
#| eval: false
#' Compute NDVI, NDRE, and GNDVI from standard band names.
#'
#' @param ms Multispectral raster with Blue/Green/Red/RedEdge/NIR.
#' @return named list of rasters.
calculate_vis <- function(ms) {
  r <- ms[["Red"]]
  n <- ms[["NIR"]]
  g <- ms[["Green"]]
  re <- ms[["RedEdge"]]
  list(
    NDVI = (n - r) / (n + r),
    NDRE = (n - re) / (n + re),
    GNDVI = (n - g) / (n + g)
  )
}
```

## 6.2 Terrain Derivatives

```{r}
#| label: fn-terrain-derivatives
#| eval: false
#' Compute slope/aspect/TPI/TWI from DEM.
#'
#' @param dem DEM raster (terra SpatRaster recommended).
#' @return named list of terrain rasters.
terrain_derivatives <- function(dem) {
  d <- terra::rast(dem)
  slope <- terra::terrain(d, v = "slope", unit = "degrees")
  aspect <- terra::terrain(d, v = "aspect", unit = "degrees")
  tpi <- terra::terrain(d, v = "TPI")
  twi <- terra::terrain(d, v = "TRI") # placeholder; replace with your preferred TWI algorithm.
  list(Slope = slope, Aspect = aspect, TPI = tpi, TWI = twi)
}
```

## 6.3 Build Covariate Stack (Retrieve Path, No Upload)

```{r}
#| label: fn-build-covariates-retrieve
#| eval: false
#' Build a covariate stack from retrieved Sentinel + elevation.
#'
#' @param sentinel_rast Sentinel SpatRaster with Blue/Green/Red/RedEdge/NIR names.
#' @param dem DEM SpatRaster.
#' @return SpatRaster containing base bands, VIs, and terrain layers.
build_covariates_retrieve <- function(sentinel_rast, dem) {
  s <- terra::rast(sentinel_rast)
  d <- terra::rast(dem)
  d <- terra::resample(d, s[[1]], method = "bilinear")

  vis <- calculate_vis(s)
  vis_r <- terra::rast(vis)
  terr <- terrain_derivatives(d)
  terr_r <- terra::rast(terr)

  out <- terra::c(s, vis_r, terr_r)
  names(out) <- make.names(names(out), unique = TRUE)
  out
}
```

## 6.4 Visual Helpers (Shiny-Like Map Feedback)

```{r}
#| label: fn-plot-aoi-and-samples
#| eval: false
#' Plot AOI, optional covariate layer, and sample points.
#'
#' @param aoi sf AOI polygon.
#' @param samples Optional sf POINT object.
#' @param covariate Optional SpatRaster/RasterLayer for background.
#' @return mapview object.
plot_workflow_map <- function(aoi, samples = NULL, covariate = NULL) {
  aoi_v <- mapview::mapview(aoi, alpha.regions = 0.15, color = "yellow", lwd = 2)
  cov_v <- if (!is.null(covariate)) mapview::mapview(covariate, layer.name = "Covariate") else NULL
  smp_v <- if (!is.null(samples)) mapview::mapview(samples, col.regions = "red", cex = 5, layer.name = "Samples") else NULL
  Reduce(`+`, Filter(Negate(is.null), list(cov_v, aoi_v, smp_v)))
}
```

```{r}
#| label: fn-plot-auto-comparison
#| eval: false
#' Visual comparison of random, cLHS, and grid sampling.
#'
#' @param aoi sf AOI polygon.
#' @param random_pts,clhs_pts,grid_pts sf POINT objects.
#' @return mapview paneled map.
plot_sampling_comparison <- function(aoi, random_pts, clhs_pts, grid_pts) {
  p1 <- mapview::mapview(aoi, alpha.regions = 0.1) + mapview::mapview(random_pts, col.regions = "deepskyblue", layer.name = "Random")
  p2 <- mapview::mapview(aoi, alpha.regions = 0.1) + mapview::mapview(clhs_pts, col.regions = "orange", layer.name = "cLHS")
  p3 <- mapview::mapview(aoi, alpha.regions = 0.1) + mapview::mapview(grid_pts, col.regions = "magenta", layer.name = "Grid")
  p1 | p2 | p3
}
```

## 6.5 cLHS Similarity Zones (Mahalanobis, Weak-GPS Friendly)

```{r}
#| label: fn-mahalanobis-distance-raster
#| eval: false
#' Compute Mahalanobis distance raster from cLHS sample distribution.
#'
#' @param covariates SpatRaster of covariates used for cLHS.
#' @param clhs_points sf POINT object returned by generate_samples(..., method = "clhs").
#' @return SpatRaster with Mahalanobis distance per pixel.
compute_mahalanobis_distance_raster <- function(covariates, clhs_points) {
  cov <- terra::rast(covariates)
  pts <- sf::st_as_sf(clhs_points)
  pts <- sf::st_transform(pts, sf::st_crs(terra::crs(cov)))

  sample_vals <- terra::extract(cov, terra::vect(pts), ID = FALSE)
  sample_vals <- stats::na.omit(as.data.frame(sample_vals))
  if (nrow(sample_vals) < 3) stop("Need at least 3 valid cLHS points to estimate covariance.")

  mu <- colMeans(sample_vals)
  S <- stats::cov(sample_vals)
  S_inv <- solve(S)

  f <- function(...) {
    m <- cbind(...)
    out <- rep(NA_real_, nrow(m))
    ok <- stats::complete.cases(m)
    if (any(ok)) {
      x <- sweep(m[ok, , drop = FALSE], 2, mu, "-")
      out[ok] <- sqrt(rowSums((x %*% S_inv) * x))
    }
    out
  }
  terra::app(cov, fun = f)
}
```

```{r}
#| label: fn-threshold-and-hist
#| eval: false
#' Choose threshold from distance quantile and visualize histogram.
#'
#' @param dist_rast Mahalanobis distance SpatRaster.
#' @param quantile_cut Quantile for representative zone (default 0.85).
#' @return Numeric threshold.
choose_similarity_threshold <- function(dist_rast, quantile_cut = 0.85) {
  v <- terra::values(dist_rast, na.rm = TRUE)
  stats::quantile(v, probs = quantile_cut, na.rm = TRUE)[[1]]
}

#' Plot distance histogram with threshold marker.
#'
#' @param dist_rast Mahalanobis distance SpatRaster.
#' @param threshold Numeric threshold.
#' @return NULL.
plot_distance_histogram <- function(dist_rast, threshold) {
  v <- terra::values(dist_rast, na.rm = TRUE)
  hist(v, breaks = 60, col = "grey80", border = "white",
       main = "Mahalanobis Distance to cLHS Feature Space",
       xlab = "Distance (lower = more similar)")
  abline(v = threshold, col = "red", lwd = 2, lty = 2)
}
```

```{r}
#| label: fn-similarity-and-weak-gps-zones
#| eval: false
#' Create similarity zone and weak-GPS recommendation zone.
#'
#' @param covariates Covariate SpatRaster.
#' @param clhs_points cLHS sample points.
#' @param quantile_cut Quantile for similarity zone.
#' @param gps_tolerance_m GPS tolerance buffer in meters.
#' @return list(distance, threshold, similarity_zone, weak_gps_zone).
create_similarity_guidance <- function(covariates, clhs_points, quantile_cut = 0.85, gps_tolerance_m = 30) {
  d <- compute_mahalanobis_distance_raster(covariates, clhs_points)
  thr <- choose_similarity_threshold(d, quantile_cut = quantile_cut)
  sim <- d <= thr
  names(sim) <- "similar_zone"

  zone_poly <- terra::as.polygons(sim, dissolve = TRUE, values = TRUE)
  zone_poly <- sf::st_as_sf(zone_poly)
  zone_poly <- zone_poly[zone_poly$similar_zone == 1, , drop = FALSE]

  pts <- sf::st_as_sf(clhs_points)
  pts_m <- sf::st_transform(pts, 3857)
  buf <- sf::st_buffer(pts_m, dist = gps_tolerance_m)
  zone_m <- sf::st_transform(zone_poly, 3857)
  weak <- suppressWarnings(sf::st_intersection(zone_m, sf::st_union(buf)))
  weak <- sf::st_transform(weak, sf::st_crs(pts))

  list(distance = d, threshold = thr, similarity_zone = sim, weak_gps_zone = weak)
}

#' Visualize similarity guidance zones.
#'
#' @param aoi AOI polygon.
#' @param clhs_points cLHS sample points.
#' @param similarity_zone Similarity zone raster.
#' @param weak_gps_zone Weak GPS recommendation polygons.
#' @return mapview object.
plot_similarity_guidance <- function(aoi, clhs_points, similarity_zone, weak_gps_zone) {
  mapview::mapview(aoi, alpha.regions = 0.1, color = "white") +
    mapview::mapview(similarity_zone, layer.name = "Similarity Zone") +
    mapview::mapview(clhs_points, col.regions = "orange", cex = 6, layer.name = "cLHS Points") +
    mapview::mapview(weak_gps_zone, alpha.regions = 0.35, color = "cyan", layer.name = "Weak GPS Recommendation")
}
```

# 7) End-to-End Workflow Helpers

## 7.1 Retrieve/Create Workflow (No Upload, Full Path to Sampling)

```{r}
#| label: cmd-create-to-sampling-workflow
#| eval: false
# Option A: create AOI visually on map (recommended)
aoi <- create_boundary_interactive()

# Option B: create AOI from coordinates (no upload)
# aoi <- create_boundary_bbox(
#   xmin = -83.40, ymin = 33.90,
#   xmax = -83.30, ymax = 34.00
# )

# Option B: upload/read AOI
# aoi <- read_boundary("path/to/boundary.geojson")

scene_list <- search_sentinel(aoi, date_start = "2024-04-01", date_end = "2024-05-01", cloud_max = 15, limit = 25)

# Retrieve Sentinel (single or median)
sentinel <- retrieve_sentinel_median(scene_list, aoi, n_use = 8)
# sentinel <- retrieve_sentinel_single(scene_list[[1]], aoi)

# Retrieve elevation and derive covariates
dem <- download_elevation(aoi, z = 12)
covariates <- build_covariates_retrieve(sentinel$rast, dem)

# Generate sampling points
samples_random <- generate_samples(aoi, covariates, method = "random", n_points = 30)
samples_clhs <- generate_samples(aoi, covariates, method = "clhs", n_points = 30)
samples_grid <- generate_samples(aoi, covariates, method = "grid", n_points = 30)

# Visual QA (like Shiny map feedback)
plot_workflow_map(aoi, samples_clhs, covariates[[1]])
plot_sampling_comparison(aoi, samples_random, samples_clhs, samples_grid)

# Similarity guidance for weak GPS farmers
guide <- create_similarity_guidance(covariates, samples_clhs, quantile_cut = 0.85, gps_tolerance_m = 30)
plot_distance_histogram(guide$distance, guide$threshold)
plot_similarity_guidance(aoi, samples_clhs, guide$similarity_zone, guide$weak_gps_zone)
```

## 7.2 Upload Workflow (Boundary -> Upload Bands -> Compute VI)

```{r}
#| label: cmd-upload-workflow
#| eval: false
aoi <- read_boundary("path/to/boundary.geojson")
ms <- stack_individual_bands(
  blue = "blue.tif",
  green = "green.tif",
  red = "red.tif",
  rededge = "rededge.tif",
  nir = "nir.tif"
)
vis <- calculate_vis(ms)
manual_pts <- manual_sampling_interactive(aoi)
plot_workflow_map(aoi, manual_pts, ms[[1]])
```

## 7.3 Mixed Workflow (Upload If Available, Otherwise Create/Retrieve)

```{r}
#| label: cmd-mixed-workflow-helper
#| eval: false
#' Run complete workflow to sampling with upload fallback.
#'
#' @param aoi Either sf AOI or NULL.
#' @param boundary_path Optional boundary file path if aoi is NULL.
#' @param sentinel_tif Optional uploaded Sentinel stack path.
#' @param date_start,date_end Used when sentinel_tif is NULL.
#' @param cloud_max,max_results Sentinel search params.
#' @param n_scenes_for_median Number of scenes in median retrieval mode.
#' @param sample_method "random", "clhs", or "grid".
#' @param n_points Number of sample points.
#' @return list(aoi, covariates, samples).
run_workflow_to_sampling <- function(
    aoi = NULL,
    boundary_path = NULL,
    sentinel_tif = NULL,
    date_start = "2024-04-01",
    date_end = "2024-05-01",
    cloud_max = 15,
    max_results = 25,
    n_scenes_for_median = 8,
    sample_method = "clhs",
    n_points = 30
) {
  if (is.null(aoi)) {
    if (!is.null(boundary_path)) {
      aoi <- read_boundary(boundary_path)
    } else {
      stop("Provide either `aoi` or `boundary_path`.")
    }
  }

  sentinel_obj <- if (!is.null(sentinel_tif)) {
    list(rast = terra::rast(sentinel_tif))
  } else {
    feats <- search_sentinel(aoi, date_start, date_end, cloud_max = cloud_max, limit = max_results)
    retrieve_sentinel_median(feats, aoi, n_use = n_scenes_for_median)
  }

  dem <- download_elevation(aoi, z = 12)
  covariates <- build_covariates_retrieve(sentinel_obj$rast, dem)
  samples <- generate_samples(aoi, covariates, method = sample_method, n_points = n_points)
  list(aoi = aoi, covariates = covariates, samples = samples)
}
```

# 8) How to Turn This Into an R Package

```{r}
#| label: package-scaffold-commands
#| eval: false
# install.packages(c("usethis", "roxygen2", "devtools", "pkgdown", "testthat"))

usethis::create_package("sentinelworkflow")
usethis::use_mit_license()
usethis::use_roxygen_md()
usethis::use_testthat()
usethis::use_package("sf")
usethis::use_package("terra")
usethis::use_package("raster")
usethis::use_package("rstac")
usethis::use_package("dplyr")
```

Suggested structure:

- `R/boundary.R` -> `read_boundary()`, `boundary_area_check()`
- `R/imagery_sentinel.R` -> search/retrieve functions
- `R/imagery_upload.R` -> upload stack/bands/VI
- `R/elevation.R` -> download/upload/derivatives
- `R/sampling.R` -> manual/generate sample APIs
- `R/similarity_zones.R` -> Mahalanobis distance and weak-GPS guidance zones
- `R/utils_indices.R` -> VI and helper math

# 9) Notes for Your Next Iteration

- First migrate **Sentinel search + single retrieve** into package functions (highest value, lowest risk).
- Keep Shiny app as orchestration/UI only.
- Add tests with one tiny AOI and very small scene count.
- Keep async/progress UI in app; keep package functions pure and deterministic.

